---
title: "R-Markdown-Wordpress-test"
author: "John Yuill"
date: "February 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(googledrive)
```

## R Markdown to Wordpress

Uses:

* library(RWordPress)
* library(knitr) options 

to upload html output from R Markdown to a Wordpress blog.

## References

Key references include:

* Yihui Xie's blog post (developer of knitr pkg, incl. knit2wp function) <https://yihui.name/knitr/demo/wordpress/>
* Details on additional options: <http://3.14a.ch/archives/2015/03/08/how-to-publish-with-r-markdown-in-wordpress/>

<strong>Note:</strong> most online resources refer to Word**P**ressLogin and Word**P**ressURL (capital **'P'**) in options but I could only get it to work with **lowercase 'p'** as in Yihui's example. 


## Example

```{r cars}
summary(cars)
```

## Will Upload Plots - but tricky

You can also upload embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

I used the code from the [detailed blog](http://3.14a.ch/archives/2015/03/08/how-to-publish-with-r-markdown-in-wordpress/) referenced above:

'To Upload plots: set knitr options'

```{r echo=TRUE, eval=FALSE}
opts_knit$set(upload.fun = function(file){library(RWordPress);uploadFile(file)$url;})
```

This worked but with warnings about deprecated functions - may not be sustainable.

## Alternative handling of graphics

Idea was to upload to Google Drive, use link to file. Doesn't work because Wordpress doesn't recognize that the link is an image file. :P

```{r longley-gnp}
#str(longley)
ggplot(longley, aes(x=Year, y=GNP))+geom_line()
## ggsave automatically saves previous plot by default (can be managed)
ggsave(filename='longley-gnp.png', device='png', path='figures')

## upload file to GDrive
#drive_upload(media='figures/longley-gnp.png', #path=as_id('1514PZY5fZglqPadDLAc8Ug6cJTsct8Ta'))

```

<file link here>